Value-Suppressing Uncertainty Maps

### 2017-03-27 paper/template.tex
Study code to prep for merge

In for visualizations like choropleth maps, heatmaps, or treemaps, visual channels such as position and length are reserved for data variables other than value (such as geographic location or relative size). In these situations, data value is habitually encoded using color. Colors in such univariate quantitative color maps should be sufficiently far apart as to be perceptually distinguishable~\cite{ware1988color}, and vary in lightness as well as hue in order to afford an implicit ordering of value~\cite{borland2007rainbow,rogowitz2001blair}. Given these constraints, and the perceptual non-orthogonality of color channels, the construction of \emph{bivariate} color maps is difficult. Viewers need to be able to independently assess the variables of interest, but the resulting bivariate map should also be expressive and easy to interpret. Bivariate maps with these constraints are often limited to relatively few categories (say, a 3x3 matrix as in Figure \ref{fig:example}) \cite{robertson1986generation,trumbo1981theory}. However, even simple bivariate maps can be difficult for the general audience to interpret \cite{wainer1980empirical}.

@@

\subsection{}

@@

If both of these assumptions hold, then it follows that the designer of a bivariate map should allocate more mark types to certain values, and fewer mark types to uncertain values. VSUMs codify this decision by \emph{reducing the number of mark categories for representing value as uncertainty increases.} The general VSUM algorithm is as follows:

@@

Correll et al. use a version of VSUMs in their LayerCake genomics visualization tool \cite{correll2015layercake,correll2011visualizing}, where marks representing genomic data with increasing uncertainty are mapped to a smaller and smaller set of increasingly grey colors, creating the effect of uncertain values retreating into ``fog'' while highly certain values remain prominent. Other bivariate visualizations implicitly alias together uncertain values. For instance, if uncertainty is encoded by transparency, a maximally uncertain glyph may be entirely transparent, and so impossible to distinguish from any other maximally uncertain glyph. Other channels with a semiotic connection ot uncertainty, such as saturation, value, blur, or size, also have deleterious effects on the disambiguation of colors and shapes. In both of these cases, this property of aliasing is ad hoc, and places no guarantees on the discriminability of colors. The binning and degradation approach of VSUMs makes this choice to alias values explicit to both the designer and the viewer, and results in a bivariate mapping with known perceptual properties.

>>>>

In for visualizations like choropleth maps, heatmaps, or treemaps, visual channels such as position and length are reserved for data variables other than value (such as geographic location or relative size). In these situations, data value is habitually encoded using color. Colors in such univariate quantitative color maps should be sufficiently far apart as to be perceptually distinguishable~\cite{ware1988color}, and vary in lightness as well as hue in order to afford an implicit ordering of value~\cite{borland2007rainbow,rogowitz2001blair}. Different choices of color maps can highlight different features of the data, and should be chosen with care based on the data types and important features to be visualized~\cite{rogowitz1996not}. Given these constraints, and the perceptual non-orthogonality of color channels, the construction of \emph{bivariate} color maps is difficult. Viewers need to be able to independently assess the variables of interest, but the resulting bivariate map should also be expressive and easy to interpret. Bivariate maps with these constraints are often limited to relatively few categories (say, a 3x3 matrix as in Figure \ref{fig:example}) \cite{robertson1986generation,trumbo1981theory}. However, even simple bivariate maps can be difficult for the general audience to interpret \cite{wainer1980empirical}.

@@

@@

If both of these assumptions hold, then it follows that the designer of a bivariate map should allocate more mark types to certain values, and fewer mark types to uncertain values. VSUMs codify this decision by \emph{reducing the number of mark categories for representing value as uncertainty increases.} This means that data encoded using a VSUM will make visible only the largest of differences in uncertain data, but highlight comparatively small differences in value when uncertainty is low. VSUMs therefore act as both a filtering mechanism (in that values with too much uncertainty are all mapped to the same glyph), as well as an implicit test of effect size (in that smaller and smaller changes in value are visible as uncertainty decreases). This strategy of dampening the low quality or high uncertainty values in maps in order to focus on more informative regions has measurable benefits, including the removal of statistically spurious visual patterns, and the highlighting of regions of interest~\cite{correll2017surprise}.

The general VSUM algorithm is as follows:

@@

Correll et al. use a version of VSUMs in their LayerCake genomics visualization tool \cite{correll2015layercake,correll2011visualizing}, where marks representing genomic data with increasing uncertainty are mapped to a smaller and smaller set of increasingly grey colors, creating the effect of uncertain values retreating into ``fog'' while highly certain values remain prominent. Other bivariate visualizations implicitly alias together uncertain values. For instance, if uncertainty is encoded by transparency, a maximally uncertain glyph may be entirely transparent, and so impossible to distinguish from any other maximally uncertain glyph. Other channels with a semiotic connection to uncertainty, such as saturation, value, blur, or size, also have deleterious effects on the disambiguation of colors and shapes. In both of these cases, this property of aliasing is ad hoc, and places no guarantees on the discriminability of colors. The binning and degradation approach of VSUMs makes this choice to alias values explicit to both the designer and the viewer, and results in a bivariate mapping with known perceptual properties.
